---
title: "Fisher's Exact Test"
draft: false
source: true
output: binb::metropolis
fontsize: 12pt
author: Max Turgeon
institute: STAT 3150---Statistical Computing
header-includes:
  - \usefonttheme{professionalfonts}
  - \usepackage{graphicx}
  - \usepackage{tikzpagenodes}
  - \usetikzlibrary{calc}
  - \usepackage{caption}
---

```{r,setup, include=FALSE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})

knitr::opts_chunk$set(cache = FALSE, message = FALSE,
                      linewidth = 50)
```

## Lecture Objectives

  - Use Fisher's Exact Tests to test for independence of two categorical variables.
  - Understand the connection between Fisher's Exact Test and permutation tests.

## Motivation {.allowframebreaks}

  - *The Lady Tasting Tea: How Statistics Revolutionized Science in the Twentieth Century* by David Salsburg
  - A lady in England claimed she could always tell whether milk or tea was poured in a cup first.
  - Fisher designed an experiment to test this claim:
    + He randomly filled four cups with tea first, and four cups with milk first.
    + He then asked the lady to guess which four cups were filled with milk first.
    
\vspace{1cm}

  - Here are the results, summarised in a table:
  
|             | Guess Milk | Guess Tea |
|-------------|:----------:|:---------:|
| Actual Milk |      3     |     1     |
| Actual Tea  |      1     |     3     |

  - Fisher's null hypothesis was that the lady was guessing randomly. 
  - Under this null hypothesis, the value in the top left cell (i.e. correct milk guesses) follows a *hypergeometric distribution*.
    + We can therefore compute the probability she correctly guessed 3.

```{r echo = TRUE, eval = TRUE}
dhyper(3, m = 4, n = 4, k = 4)
```
   
   - **What would be a more extreme value?** 
     + If she had guessed 4 correctly! 
   - The p-value is the sum of the probability mass function at $x = 3$ and $x = 4$.
   
```{r echo = TRUE, eval = TRUE}
dhyper(3, m = 4, n = 4, k = 4) +
  dhyper(4, m = 4, n = 4, k = 4)
```

  - Given that probability, Fisher concluded there was a lack of evidence that the lady could really tell which was poured first.
  
## Fisher's Exact Test {.allowframebreaks}

  - Fisher's exact test is a test of independence between two factors 
    + E.g. the true liquid poured first and the lady's guess.   - It relies on an important assumption:
    + Both the **row totals** and the **column totals** should be *fixed* by design. 
  - This was the case in Fisher's experiment: he poured 4 cups of each, and told the lady, so each row and each column in the table above has to sum to 4.
  
\vspace{1cm}

  - This assumption is key: 
    + It implies that the value of a single cell completely determines the value of the other three cells. 
  - Therefore, any of the four cells is a valid test statistic. 
  - E.g. if we pick the lower left cell (i.e. wrong milk guesses), we get the same p-value (but note the difference in which values are more extreme!):

```{r echo = TRUE, eval = TRUE}
dhyper(1, m = 4, n = 4, k = 4) +
  dhyper(0, m = 4, n = 4, k = 4)
```

## Permutation tests

  - **What is the connection with permutation tests?** 
  - Under the null hypothesis, the value of any cell follows a hypergeometric distribution. 
  - We can quantify the likelihood that a permutation would lead to a valid 2x2 table (i.e. with the correct row and column sums).
  - As we permute the data, the frequency of each configuration converges to the hypergeometric probabilities.

## Example {.allowframebreaks}

```{r}
data <- rep(c("Milk", "Tea"), each = 4)
data
```


```{r}
results <- replicate(1000, {
  data_perm <- sample(data)
  # Treat first 4 values as if the lady 
  # had guessed milk
  sum(data_perm[1:4] == "Milk")
})

mean(c(3, results) >= 3)
```

  - As we can see, we get a similar p-value.

## Summary

  - Fisher's Exact Test is very much in the same spirit as the permutation tests.
  - But computational resources were limited.
    + Fisher couldn't *actually* permute the data.
    + Instead, he had to compute the probability of each permutation.
  - **Note**: you can construct a valid permutation test that doesn't rely on the assumption of fixed margins.