---
title: "`R` preliminaries"
draft: false
source: true
output: binb::metropolis
fontsize: 12pt
author: Max Turgeon
institute: STAT 3150--Statistical Computing
header-includes:
  - \usefonttheme{professionalfonts}
  - \usepackage{graphicx}
  - \usepackage{tikzpagenodes}
  - \usetikzlibrary{calc}
  - \usepackage{caption}
---

```{r,setup, include=FALSE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})

knitr::opts_chunk$set(cache = FALSE, message = TRUE,
                      linewidth = 50)

set.seed(3150)
```

# Quick Introduction

## Why use `R`?

  - Modern statistics relies heavily on statistical computing.
    + Simulation studies
    + Data analysis
  - `R` is a programming language that can be used on most platforms (Mac, Windows, Linux, Solaris, etc.)
  - `R` is very flexible.
    + It can be extended via `R` packages.
  - `R` offers a powerful interface for analyzing data and producing high-quality plots.
    + Extensive ecosystem of packages (unlike Julia)

## Interacting with `R`

  - You can interact with `R` in many ways:
    + Through the command line
    + In batch mode (i.e. running a script)
    + Through an Integrated Development Interface (IDE)
  - I strongly recommend using RStudio, which is the most powerful IDE for `R`.
  
## `R` is a **functional** programming language

  - In `R`, functions are *first-class citizens*:
    + They can be assigned to variables
    + They can be passed as function arguments
    + They can be returned by other functions
  - Everything that happens in `R` is a function call.
    + E.g. Control structures are implemented as functions too!
  - Therefore, to become effective in `R`, learn to write functions

## Every `R` object is a vector

  - Even scalars are vectors of length 1
  - There are two main types of vectors:
    + **Atomic vectors**: each element is of the same primitive type (e.g. numeric, boolean, character)
    + **Lists**: elements can be of any type, even lists!
  - Matrices and arrays are also vectors, but with extra structure.
    
\begin{center}
A very common pattern in \texttt{R} is to apply functions to vectors (as opposed to using \texttt{for} loops). A function that takes a vector as input is called \textbf{vectorized}.
\end{center}

## Example {.allowframebreaks}

```{r}
# Create a vector
vect <- c(2, 6, 3, 5.5)

# What is its mean?
# Using for loops
n <- length(vect)
sum <- 0 # Initialize

for (i in 1:n) { # R is 1-indexed!
  sum <- sum + vect[i]
}
```

```{r}
mean <- sum/n
mean
```

```{r}
# In R, use vectorized functions
# whenever possible
mean(vect)
```

# Main object types

## Variables

  - Variables are ways to assign values (or objects) to names (or symbols).
    + E.g. `vect <- c(1,2,3,4)`
  - This allows us to write more robust and flexible code.
  - Use meaningful names to make code *human-readable*. 
    + Try to use `n.sample` or `sample_size` or `createMatrix`, instead of `n` or `nn`.
    + Have a look at the Tidyverse style guide for `R`: https://style.tidyverse.org/
  - Descriptive names make it easier to design, debug, and improve your code.


## Assignment Operator

  - The are several ways of assigning a value to a variable.

```{r}
# These are all equivalent
x <- 10
10 -> x
x = 10
assign("x", 10)
```

  - For readability of the code, the preferred option is `<-`. 
    + Although `=` also works, it is usually reserved for function arguments.

## Atomic vectors {.allowframebreaks}

  - Recall: An atomic **vector** is a sequence of values, all of the same primitive type.

```{r}
x <- c(0, 5, 12, 8)
x
```

  - The `c` function (for *concatenate*) returns a vector made from all the given arguments.

```{r}
y <- c(3, 2)
c(x, y)
```

  - If elements are not all of same type, `R` tries to coerce them.
  
```{r}
c(1, 2.5, "Stat", FALSE)
```


  - `R` has a built-in function to create sequences.

```{r}
seq(from = 1, to = 3, by = 0.5)
# Equivalently
seq(1, 3, by = 0.5)
```

```{r}
# Decreasing sequences
seq(1, 0, by = -0.2)
```

  - There is also a shorthand for sequences of consecutive integers:

```{r}
1:5
1:(-2)
```

\vspace{1cm}

  - Accessing one element of a vector:

```{r}
x[2]
```

 - Accessing more than one element:
 
```{r}
 x[c(1, 3)]
```

  - Accessing all but some elements:

```{r}
x[c(-2, -4)]
```

  - Accessing can also be done with a boolean vector:

```{r}
x_large <- x > 7
x_large
x[x_large]
```

  - Or using the `which` function (returns the *indices* of the elements of a boolean vector that are `TRUE`)

```{r}
which(x_large)
x[which(x_large)]
```

## Comparisons and Logical Operators {.allowframebreaks}

  - Comparisons are made like most other languages:

```{r}
7 <= 5
7 != 5
```

\vspace{1in}

```{r}
# Even works with character values
"abc" < "bca"
```

  - Recall that `=` is an assignment operator. Equality is checked with a double equal sign:

```{r}
7 == 5
```

  - For vectors of length > 1, comparisons are actually done component-wise:

```{r}
y <- rep(10, times = 4) # rep for repeat
y
x < y
```

  - Because of *recycling*, this is equivalent to:

```{r}
x < 10
```

  - The basic logical operators are `|` (or) and `&` (and). They also work component-wise:

```{r}
(x > 3) & (x < 10)
(x < 3) | (x > 10)
```

## Vector arithmetic

  - Arithmetic operators work component-wise:

```{r}
z <- 1:4
x + z
x * z
z / x
```

## Recycling

  - Binary operators (arithmetic, comparison, logical) are applied element-wise to vectors.
  - `R` uses the concept of **recycling** when applying these operators to vectors of different lengths:
    + repeat the shorter vector enough times to obtain a new vector of the same length as the longer vector
    + apply the operator to the two longer vectors thus obtained
    + if the length of the longer vector is not a multiple of the length of the shorter vector, `R` returns a warning.

## Arrays and Matrices {.allowframebreaks}
  
  - **Arrays** are tables made from elements of the same type, like atomic vectors.
  - You can create arrays from atomic vectors by specifying the dimensions.
    + **Note**: `R` is column-major, which means it fills the matrix column by column (instead of by row)

```{r}
A_mat <- matrix(1:4, nrow = 2, ncol = 2)
A_mat
```

  - If you prefer filling by row:

```{r}
B_mat <- matrix(1:4, nrow = 2, ncol = 2, byrow = TRUE)
B_mat
```

  - Array arithmetic (`+`, `*`, etc.) is done component-wise:
  
```{r}
A_mat * B_mat
```


## Accessing elements of arrays {.allowframebreaks}

  - Accessing elements of arrays can be done by proper indexing of the array itself or by indexing the underlying vector:

```{r}
A_mat[1, 2]
A_mat[3]
```

- You can also select a full row or columns:

```{r}
B_mat[, 2]
```

## Basic matrix operations {.allowframebreaks}

  - Some basic operations: `t`, `det`, `%*%`.  
  - Matrix multiplication requires the dimension of the involved matrices to match.  

```{r}
B_mat %*% t(A_mat)
```
 
  - `R` treats vectors as column-vectors or row-vectors, as needed.

```{r}
A_mat %*% c(2, 3)
c(5, 1) %*% A_mat
```

  - Matrix inversion is done with `solve`:

```{r}
solve(A_mat)
```

## Lists {.allowframebreaks}

  - Unlike atomic vectors, **lists** are sequences of values, not necessarily all of the same type.

```{r}
course <- list(3150, "Statistical Computing", 
               FALSE, 3.0)
course
```

  - Lists are vectors, and they can be subsetted using `[ ]`.

```{r}
course[4]
```

  - **Note**: the above is still a list! To extract the element, use double brackets:

```{r}
course[[4]]
```

  - Use `c` to add elements to a list (just like atomic vectors).

```{r}
c("STAT", course)
```

## Names for lists {.allowframebreaks}

  - Very often, elements of a list will be given names.

```{r}
names(course) <- c("Number", "Title", 
                   "Lab", "Credit_hours")
course
```

  - There is a shortcut for using names with lists:

```{r}
course[["Title"]]
course$Title
```

  - Lists can be created with names.

```{r}
other_course <- list(Number = 4150, 
                     Title = "Bayesian Statistics", 
                     Lab = FALSE, Credit_hours = 3.0)
other_course
```

- A named element can be added to a list.

```{r}
course$Dept <- "STAT"
course
other_course[["Dept"]] <- "STAT"
```

## Data Frames {.allowframebreaks}

  - A **data frame** is a list of vectors that are all of the same length. Importantly, the vectors *can be of different types*.  
  - Data frames are how `R` models datasets:  
    + Columns are variables,  
    + Rows are units or subjects.  

```{r}
courses <- data.frame(Dept = c("STAT", "STAT"), 
                      Number = c(3150, 4150), 
                      Title =c("Statistical Computing",
                                "Bayesian Statistics"),
                      Lab = c(FALSE, FALSE), 
                      Credit_hours = c(3.0, 3.0))
```


```{r}
courses
```

  - Elements of data frames can be accessed, like matrices, by indices or names.

```{r}
courses[2, "Title"]
```

  - The shortcut `$` works with the columns of data frames:

```{r}
courses$Lab
```

# Control Structures

## Conditional Statements {.allowframebreaks}

- The function **if** is used to control which of two blocks of code are executed.  
- The typical syntax is:

```{r, eval=FALSE}
if (condition) {
  # Block of code to be executed 
  # when condition is TRUE
} else {
  # Another block of code to be executed 
  # when condition is FALSE
}
```

 - Braces are not necessary when a block contains only one line of code, but it is good practice to use the above syntax.  
 - The `else` statement is not required.
 
## Example 

```{r}
# Sample 1 value from a standard normal
x <- rnorm(1)
if (x < 0) {
  message("The observation x is negative.")
} else {
  message("The observation x is positive.")
}
```

## Conditional Statements (cont'd) {.allowframebreaks}

  - It is also possible to have more than one else statement:

```{r}
x <- rnorm(10)
loc_measure <- "mid_point"
```


```{r}
if (loc_measure == "mean") {
  mean(x)
} else if (loc_measure == "median") {
  median(x)
} else if (loc_measure == "mid_point") {
  0.5*(min(x) + max(x))
} else {
  stop(paste("You have to choose between mean,",
             "median and mid_point."))
}
```

  - The above is referred to as a *nested if* structure.
  - The `switch` function can also be used in the above setting:

```{r, error = TRUE}
loc_measure <- "truncated_mean"
switch(loc_measure,
       "mean" = mean(x),
       "median" = median(x),
       "mid_point" = 0.5*(min(x) + max(x)),
       stop(paste("You have to choose between mean,\n",
             "median and mid_point."), call. = FALSE)
       )
```

## `for` Loop {.allowframebreaks}

  - The `for` statement specifies that a certain operation should be repeated a *fixed* number of times.  
  - The syntax is:

```{r, eval = FALSE}
for (element in vector) {
  # Block of code to be repeated
  # once for each element of vector
}
```

## Example

```{r}
for (k in 3:0) {
  message(k)
  if (k == 0) message("Blast off!")
}
```

## Exercise

Approximate a geometric sum using a finite number $N$ of terms:

$$S = \sum_{k=0}^{\infty} \left( \frac{1}{2} \right)^k \approx \sum_{k=0}^{N-1} \left( \frac{1}{2} \right)^k.$$

## Solution {.allowframebreaks}

```{r}
N <- 10
approx <- 0
for (k in 0:(N - 1)) {
  # update current approx by adding next term
  approx <- approx + 2^{-k}     
}
approx
```

  - **Note**: it is more efficient to use vectorized functions:

```{r}
sum(0.5^(0:(N-1)))
```

## `while` Loop

  - The *while* loop repeats an expression for as long as a condition holds.
  - The syntax is:

```{r, eval = FALSE}
while (condition) {
  # Block of code to be repeated 
  # as long as condition is TRUE
}
```

## Example {.allowframebreaks}

```{r}
num_flips <- 0
flip <- "tails"

while (flip == "tails") {
  # Flip a coin
  flip <- sample(c("tails", "heads"), size = 1)
  num_flips <- num_flips + 1
}
```

\vspace{2in}

```{r}
# How many flips?
num_flips
```

## Example {.allowframebreaks}

  - What if we want to approximate the geometric sum within $\epsilon = 10^{-7}$ of the true value of $S$?
  
```{r}
approx <- 0
current_err <- 2
k <- 0
while(current_err > 10^(-7)) {
  approx <- approx + 2^(-k)
  current_err <- 2 - approx
  k <- k + 1
}
```

```{r}
# How many terms?
k
```

## Creating functions {.allowframebreaks}

  - Creating new functions is an important part of programming.  
  - This is done with the function `function` and through assignment.

```{r, eval = FALSE}
new_function <- function(arg1, arg2 = def_val) {
  # Block of code to be executed using the arguments
  return(value)
}
```

  - This creates a function named `new_function` that can then be used like any other `R` function. 
  - The function has two arguments:
    + `arg1` is required
    + `arg2` has the default value `def_val`
  - The function will return the output of the last statement, unless it hits `return` (after which it exits).

## Example {.allowframebreaks}

  - Consider the following piecewise linear function:

$$f(x) = \begin{cases}
1 & \text{if } x<-1 \text{ or } x>1, \\
-x & \text{if } -1 \leq x <0, \\
x & \text{if } 0 \leq x \leq 1.
\end{cases}$$

  - In `R`, this can be defined as follows:
  
```{r}
fun <- function(x) {
  if ((x >= -1) & (x < 0)) {
    value <- -x
  } else if ((x >= 0) & (x <= 1)) {
    value <- x
  } else value <- 1
  
  return(value)
}
```

\vspace{1cm} 

```{r}
c(fun(-3), fun(-0.3), fun(0.4), fun(1.5))
```

  - Equivalently, we can define

```{r}
fun_cleaner <- function(x) {
  if ((x >= -1) & (x < 0)) {
    return(-x)
  }
  if ((x >= 0) & (x <= 1)) {
    return(x)
  }
  return(1)
}
```

```{r}
c(fun_cleaner(-3), fun_cleaner(-0.3), 
  fun_cleaner(0.4), fun_cleaner(1.5))
```

## Exercise

  - Implement the following, more general, function, where $a > 0$ is arbitrary:
  
$$f_a(x) = \begin{cases}
a & \text{if } x < a, \text{ or } x>a, \\
-x & \text{if } -a \leq x <0, \\
x & \text{if } 0 \leq x \leq a,
\end{cases}$$

  - *Hint*: Use a second argument to the function.